---
output: 
  html_document: 
    keep_md: yes
---
# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

First loaded needed libraries. Then unzipped the zip file using `unzip()`. 
After that loaded the data using `read.csv()`

```{r echo = TRUE}
library(ggplot2)
unzip("activity.zip")
dat <- read.csv("activity.csv")
dat$date <- as.Date(dat$date)
head(dat)
```

## What is mean total number of steps taken per day?

Found out total number of steps taken per day using `aggregate()`. Then plotted 
a histogram of number of steps taken per day using `ggplot()`

```{r echo = TRUE}
totalsteps <- aggregate(steps ~ date, dat, sum)
g1 <- ggplot(totalsteps, aes(steps))
p1 <- g1 + geom_histogram(bins = 25, fill = "orange")
print(p1)
```

Mean and meadian can be found out using `mean()` and `median()`
```{r echo = TRUE}
mean(totalsteps$steps)
median(totalsteps$steps)
```

Here mean = 10766.19 and median = 10765

## What is the average daily activity pattern?

We can find out average value of steps taken per interval by using `aggregate()`

```{r echo = TRUE}
avgsteps <- aggregate(steps ~ interval, dat, mean)
head(avgsteps)
```

Now from this data frame we can plot average value per interval as a time series
using `ggplot()`

```{r echo = TRUE}
g2 <- ggplot(avgsteps, aes(interval, steps))
p2 <- g2 + geom_line()
print(p2)
```

We can also find which interval has the maximum average steps taken 

```{r echo = TRUE}
avgsteps[which.max(avgsteps$steps),]$interval
```
This means the interval 830 to 835 has the maximum step count.

## Imputing missing values

For all the missing values, just the average step count corresponding to their 
interval is taken. Total number of missing values are

```{r echo = TRUE}
sum(is.na(dat$steps))
```

There are 2304 missing values. We create a new data frame and replace missing
values.

```{r echo = TRUE}
datnew <- dat
for (i in 1:nrow(datnew)) {
    if (is.na(datnew$steps[i])) {
        datnew$steps[i] <- avgsteps[datnew$interval[i] == avgsteps$interval,2]
            }
}
head(datnew)
```

As we can see, first few values were missing in the original data set and now
its filled. We can plot a histogram like before, this time without missing 
values

```{r echo = TRUE}
totalstepsnew <- aggregate(steps ~ date, datnew, sum)
g3 <- ggplot(totalsteps, aes(steps))
p3 <- g1 + geom_histogram(bins = 25, fill = "orange")
print(p3)
```

And can also find mean and median like before

```{r echo = TRUE}
mean(totalstepsnew$steps)
median(totalstepsnew$steps)
```

Mean is the same as the last time since we used a mean value to fill the missing
values. But median is a little different because it depends on which values were
missing.


## Are there differences in activity patterns between weekdays and weekends?

For this, first we have to create a new variable which differentiates between
weekdays and weekends. For this we can use `weekdays()` function on date 
variable and separate them accordingly

```{r echo = TRUE}
for (i in 1:nrow(datnew)) {
    if(weekdays(datnew$date[i]) == "Sunday" | 
       weekdays(datnew$date[i]) == "Saturday") {
        datnew$day[i] <- "weekend"
    }else datnew$day[i] <- "weekday"
}
head(datnew)
```
A new column is visible on the data set. Now we can use `ggplot()` to create a 
plot like this

```{r echo = TRUE}
g4 <- ggplot(datnew, aes(interval, steps))
p4 <- g4 + geom_line(stat = "summary", fun = "mean") +  facet_grid(day ~ .)
print(p4)
```
